Reducing RES Droughts through the integration of wind and PV

### 2024-11-27 Article/main.tex
Add changes to the article

%opening
\title{ \textcolor{blue}{Change Title} The Impact of Model Assumptions on RES Drought Analysis}
\author{Boris Morin, Damian Flynn, Aina Maimó Far, Conor Sweeney}

@@

As the share of renewable energy sources (RES) like wind and photovoltaic (PV) power grows, managing their variability becomes a critical challenge for energy systems. This study investigates periods of low renewable generation, known as RES droughts, focusing on Ireland's current energy mix and a projected scenario with a more balanced distribution of wind and PV capacity. Using three datasets RES droughts are analysed over 45 years to assess their frequency, duration, and return periods. The methodology involves comparing modelled capacity factors against observed data to evaluate the accuracy of each dataset and applying a fixed threshold to identify RES drought events. Results show that adding PV capacity significantly reduces the frequency and duration of RES droughts, with a more balanced energy mix minimising seasonal drought patterns. These findings underscore the importance of diversification in RES capacity to enhance energy security and resilience. \\

@@

\textcolor{blue}{Make sure past tense is used everywhere, e.g. ``was", not ``is"}

As the global shift towards renewable energy accelerates, the European Union (EU) has emerged as a leader in climate and energy policy. Under the revised European Green Deal \cite{greendeal2023report}, the EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022 \cite{europe2023stat} \textcolor{cyan}{Can't seem to find a proper reference}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources like wind and photovoltaic (PV) power. This challenge is compounded by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions~\cite{bloomfield2021, bloomfield2016, vanderWiel2019drought}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to grid stability and energy security, emphasizing the need for a resilient energy system to meet both growing electricity demand and decarbonization targets.

\textcolor{blue}{Insert sales pitch paragraph here}
\begin{comment}
This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the existing energy mix, predominantly driven by wind, while the second is based on a projected future energy scenario, featuring a more balanced share between wind and PV. A uniform methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, duration and seasonality are presented. 
\end{comment}

This study focuses on Ireland, a region with a strong reliance on wind energy and ambitious targets for PV expansion, to provide valuable insights into the potential benefits of diversifying the renewable energy mix. This work compares the performance of energy models and highlights the role of increased PV capacity in reducing RES drought risks, offering insights for policymakers and energy planners.

@@

RES droughts are identified using wind and PV CF time series. In this study, three different datasets are used, all of which are driven by ERA5 data~\cite{hersbach2020era5}. Two of the datasets are part of C3S Energy (C3S-E)~\cite{cds2023energy}, an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\cite{dubus2023energy} (C3S). One of the C3S-E datasets provides CF time series aggregated at the national scale, while the other provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset was generated using the Atlite model~\cite{hofman2021atlite}, which converts the  ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models. Atlite is an open-source tool developed by PyPSA~\cite{hofman2021atlite} and is widely used for estimating wind and PV generation~\cite{mockert2023drought, li2023atlite, parzen2023atlite, ali2023comparative}.

@@

This study uses publicly available datasets to construct and validate the models for estimating the CF of wind and PV energy. The primary data sources include: EirGrid, the transmission system operator for the island of Ireland; the ERA5 reanalysis dataset; and the C3S-E datasets.

@@

EirGrid, which operates in the Republic of Ireland, and SONI, the system operator for Northern Ireland, provide detailed datasets on all wind and PV farms across the island of Ireland (Republic of Ireland and Northern Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include information such as each farm’s installed capacity, name, and connection date. To enhance the accuracy of this data, the longitude and latitude for each farm were manually determined through online searches. For simplicity, this data will be referred to as originating from EirGrid, and the combined regions of the Republic of Ireland and Northern Ireland will be referred to as Ireland throughout the remainder of this document.

The spreadsheet available from the EirGrid website contains two key variables: generation and availability values for wind (from 2014 onward) and PV (from 2018 onward). Availability refers to the total energy generated by the RES farm, including any energy that is intentionally reduced by EirGrid or SONI to maintain grid stability. This value is combined with installed capacity to compute the CF, the main metric for assessing renewable energy output.

The PV installed capacity available on the spreadsheets from EirGrid represents the Maximum Export Capacity (MEC) and does not accurately reflect the installed PV capacity. To adjust for actual PV generation potential, installed capacities are estimated to be 1.4 times the MEC values. This installed capacity is used for PV generation, with farm output capped to the MEC value.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables which have been used by Atlite and C3S-Energy.

@@

C3S has developed a renewable energy dataset for Europe~\cite{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and PV energy from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. The time series are also available for download at two aggregated scales: regional (NUTS 2) and national.

@@

This study relies on three datasets to analyse RES droughts across the island of Ireland. Data downloaded from C3S-E are used to obtain two datasets: one based on national-level data (C3S-E N), and another on grid-level data (C3S-E G). The third dataset is computed using the Atlite model (Atlite).

@@

For national-level analyses, the aggregated CF time series provided by C3S-E are used at two levels: Republic of Ireland (NUTS0: IE) and Northern Ireland (NUTS2: UKN0). We compute a weighted average of these, based on the installed capacity of each one, to represent the total CF of Ireland. This CF time series for Ireland is based on the underlying C3S-E assumption of RES generation at every 0.25\textdegree grid point in Ireland.

@@

For grid-level analyses, the gridded dataset from C3S-E is used to create a representative time series for Ireland's RES farms. A comprehensive list of RES farms, including each farm’s GPS coordinates and installed capacity, is compiled from the EirGrid website. Using these coordinates, the nearest grid point on the C3S-E grid is identified for each farm. The CF values from the C3S-E dataset corresponding to these grid points are retrieved for each farm. To construct the final CF time series, a weighted average of the CF values is calculated, with the installed capacity of each farm serving as the weight. This process results in a single time series for each energy source (wind and PV), providing an accurate representation of RES generation in Ireland at a grid level.

@@

In addition to analysing wind and PV generation separately, a combined CF is computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). Given that PV capacity in Ireland remains relatively low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These values are based on targets outlined in the roadmap published by EirGrid and SONI~\cite{eirgrid2023future}. New time series are generated for both the Atlite and C3S-E G PV models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remains unchanged, as significant shifts in wind farm distribution are not expected. The PV and wind time series are then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

In total, twelve CF time series are analysed in this study, six for individual wind and PV CF (three models for each source) in the existing scenario, and an additional six time series that include the combined CF for both existing and balanced scenarios across the different models.

@@

In this study, a RES drought event is defined as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period of longer than 24 hours. The choice of this threshold is somewhat arbitrary but aligns with similar studies on low renewable energy production \cite{kaspar2019drought, ohba2022drought, mayer2023drought}. By using a 24-hour moving average, fewer but longer-lasting events are captured compared to using the raw CF time series, which can be more sensitive to short-term fluctuations. A fixed threshold approach has been chosen in this study to enable consistent inter-comparison between datasets.

@@

The accuracy of the datasets used in this study is verified, before continuing to the analysis of RES droughts. For the verification process, time-varying values of installed capacity is used to account for changes in RES development over the verification period. This step allows us to assess how well the datasets represent the production of renewable energy by comparing them against observed data.

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve}a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provides four associated smoothed power curves. The smoothing is done using a Gaussian filter with different standard deviations that depend on the wind speed. A separate wind CF time series for Ireland was generated for each of the wind turbine power curves and smoothing levels. The performance of each CF time series was then calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The Atlite model allows the user to select certain PV panel characteristics. In this study, the three PV panel types available are considered in the Atlite model. Following the same methodology as in the previous section, the three available models are being compared using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, the Breyer PV panel model is selected \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity, leading to uncertainties in PV generation data and the actual generating capacity in the first few months after each farm is connected. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events.

@@

In this section, RES drought events are evaluated under two different scenarios with fixed installed capacities: the existing scenario, with 89\% of wind capacity (6.7 GW) and 11\% of PV capacity (0.6 GW), and a balanced scenario, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios are driven by 45 years of ERA5 data. Using the RES drought identification process described in Section~\ref{sec:res_drought}, wind and PV droughts are first analysed separately before presenting the results for combined (wind + PV) RES droughts.

@@

For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to PV's typically lower CF compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves differently than the two C3S-E datasets. The Atlite results show a log-linear increase but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold of 0.1 CF. Atlite remains slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

For the existing installed capacity scenario (Fig.\ref{fig:return_periods}c), the return periods mirror those of wind (Fig.\ref{fig:return_periods}a), due to the low levels of installed PV. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event, shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}, extends from roughly six months for the existing scenario, to four years for the balanced scenario in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets. \textcolor{blue}{Shorten or remove: }The return periods for Atlite align more closely with those of C3S-E in the balanced scenario compared to the existing one. This convergence is due to the reduced proportion of wind capacity in the balanced scenario, which lessens the impact of wind-specific characteristics. Notably, there is a greater difference in how each model represents wind generation than PV, meaning the balanced scenario reduces these wind-related discrepancies, bringing the datasets closer together.

@@

The seasonality of RES droughts is analysed by comparing the percentage of hours in each month classified as part of a RES drought.

@@

In this study, a threshold CF value of 0.1 has been used to identify RES droughts. To assess how sensitive the results are to this threshold, the annual number of events was calculated across a range of CF threshold values (Fig.~\ref{fig:number_days_threshold}).

@@

This study has highlighted the key differences between the three datasets: Atlite, C3S-E G, and C3S-E N in their ability to model RES droughts. One of the most significant differences is how each dataset incorporates the specific locations of RES farms. Both Atlite and C3S-E G consider the locations of wind and PV farms, which should, in theory, provide a more accurate representation of RES generation. While this approach improves PV drought models, our analysis indicates that for wind, the Atlite dataset performs better overall, especially in its close alignment with EirGrid data for wind generation estimates. This finding suggests that although the inclusion of RES farm locations is beneficial, particularly for PV, the accuracy of the power curve model in Atlite may be the critical factor in producing realistic time series for low-generation events.

The results show that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. In contrast, Atlite stands out for its closer alignment with EirGrid data, particularly in wind generation, as seen in the verification process (Section \ref{sec:wind_verification}). This distinction is also evident in the analysis, where Atlite reports higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. One of the study's most critical findings is that the accuracy of RES drought modelling relies more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting accurate models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis indicates a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. This balanced approach offers a more stable and secure energy supply by mitigating RES drought risks. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.

In terms of future research, there are two main topics to explore. First, integrating future climate data with energy scenarios to better understand how climate change might affect RES droughts. Second, expanding the scope of the study to include Europe would provide a more comprehensive understanding of RES drought in an interconnected energy grid. However, this expansion would require extensive verification across all EU countries, making it a more complex but highly relevant challenge.

\newpage
\section{Acknowledgment}
Using the Climate Data Store. Thank EirGrid for the data and add a reference to the funding project or projects

>>>>

%opening
%\title{ \textcolor{blue}{Change Title} The Impact of Model Assumptions on RES Drought Analysis}
\title{Reducing RES Droughts through the integration of wind and PV}
\author{Boris Morin, Damian Flynn, Aina Maimó Far, Conor Sweeney}

@@

As the share of renewable energy sources (RES) like wind and photovoltaic (PV) power grows, their dependence on weather poses a critical challenge for energy systems. This study investigates periods of low renewable generation, known as RES droughts, focusing on Ireland's current wind-dominated energy mix and a projected scenario with a more balanced distribution of wind and PV capacity. Three different modelled datasets are used, and their capacity factors are compared to observed data to quantify their overall skill. Then, using a RES drought definition with a fixed threshold, their skill at representing these events is assessed. These modelled capacity factors allow for an expansion over a 45-year period, allowing for an analysis of the frequency, duration and return periods of RES droughts at a climatological scale. Results show the importance of model selection for RES drought risk assessment. On a similar note, the addition of PV capacity to a wind-dominated system results in a large reduction in the frequency and duration of RES droughts, with a more balanced energy mix minimising seasonal drought patterns. These findings underscore the importance of diversification in RES capacity to enhance energy security and resilience. \textcolor{magenta}{I rewrote this and I still do not like it}

% As the share of renewable energy sources (RES) like wind and photovoltaic (PV) power grows, their dependence on weather poses a critical challenge for energy systems. This study investigates periods of low renewable generation, known as RES droughts, focusing on Ireland's current wind-dominated energy mix and a projected scenario with a more balanced distribution of wind and PV capacity. Using three datasets RES droughts are analysed over 45 years to assess their frequency, duration, and return periods. The methodology involves comparing modelled capacity factors against observed data to evaluate the accuracy of each dataset and applying a fixed threshold to identify RES drought events. Results show that adding PV capacity significantly reduces the frequency and duration of RES droughts, with a more balanced energy mix minimising seasonal drought patterns. These findings underscore the importance of diversification in RES capacity to enhance energy security and resilience. \\

@@

As the global shift towards renewable energy accelerates, the European Union (EU) has emerged as a leader in climate and energy policy. Under the revised European Green Deal \cite{greendeal2023report} \url{https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=SWD%3A2022%3A230%3AFIN&qid=1653033922121}, the EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022 \cite{europe2023stat} \url{https://ec.europa.eu/eurostat/statistics-explained/index.php?title=Renewable_energy_statistics}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources like wind and photovoltaic (PV) power. This challenge is compounded by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions~\cite{bloomfield2021, bloomfield2016, vanderWiel2019drought}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to grid stability and energy security, emphasizing the need for a resilient energy system to meet both growing electricity demand and decarbonization targets.

\begin{comment}
This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the existing energy mix, predominantly driven by wind, while the second is based on a projected future energy scenario, featuring a more balanced share between wind and PV. A uniform methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, duration and seasonality are presented. 
\end{comment}

This study focuses on Ireland, a region with a strong reliance on wind energy and ambitious targets for PV expansion. This makes it the perfect case study to provide valuable insights into the potential benefits of diversifying the renewable energy mix on RES droughts. This work also compares the performance of different energy models, revealing how modelling decisions impact the representation of RES droughts. The combination of these two offers insights about the effect of system planning decisions on RES droughts and the responses required to minimise their impact.

\begin{comment}
This study focuses on Ireland, a region with a strong reliance on wind energy and ambitious targets for PV expansion, to provide valuable insights into the potential benefits of diversifying the renewable energy mix. This work compares the performance of energy models and highlights the role of increased PV capacity in reducing RES drought risks, offering insights for policymakers and energy planners.
\end{comment}

@@

RES droughts are identified using wind and PV CF time series. In this study, three different datasets are used, all of which are driven by ERA5 data~\cite{hersbach2020era5}. Two of the datasets are part of C3S Energy (C3S-E)~\cite{cds2023energy}, an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\cite{dubus2023energy}. One of the C3S-E datasets provides CF time series aggregated at the national scale, while the other provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset was generated using the Atlite model~\cite{hofman2021atlite}, which converts the  ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models. Atlite is an open-source tool developed by PyPSA~\cite{hofman2021atlite} and is widely used for estimating wind and PV generation~\cite{mockert2023drought, li2023atlite, parzen2023atlite, ali2023comparative}.

@@

This study uses publicly available datasets to construct and validate the models for estimating the CF of wind and PV energy. The primary data sources include: EirGrid, the transmission system operator (TSO) for the Republic of Ireland; the ERA5 reanalysis dataset; and the C3S-E datasets.

@@

EirGrid, the TSO for the Republic of Ireland, and SONI, the Northern Ireland TSO, provide detailed datasets on all wind and PV farms across the island of Ireland (Republic of Ireland and Northern Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include information such as each farm’s installed capacity, name, and connection date. To enhance the accuracy of this data, the longitude and latitude for each farm were manually determined through online searches. For simplicity, this data will be referred to as originating from EirGrid, as all-island data was directly obtained from EirGrid, and the combined regions of the Republic of Ireland and Northern Ireland will be referred to as Ireland throughout the remainder of this document.

The spreadsheet available from the EirGrid website contains two key variables: generation and availability. Generation is the energy that a RES farm actually contributed to the grid, which may include limitations introduced by the TSO to maintain grid stability, most commonly in the form of curtailment. Availability represents the energy that would have been generated from a RES farm if no grid constraints had been applied, making it representative of the weather-related response. This is then combined with installed capacity to compute the CF, the main variable for assessing renewable energy output. Generation and availability values are available from 2014 onward for wind and from 2018 onward for PV, although PV only became present in the Republic of Ireland data in 2023.

The PV installed capacity available on the spreadsheets from EirGrid represents the Maximum Export Capacity (MEC) and does not accurately reflect the installed PV capacity. To adjust for actual PV generation potential, installed capacities were estimated to be 1.4 times the MEC values. This installed capacity was used for PV generation, with farm output capped to the MEC value.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used by Atlite and C3S-Energy.

@@

The EU Copernicus Climate Change Service developed the C3S-E renewable energy dataset for Europe~\cite{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and PV energy from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. The time series are also available for download at two aggregated scales: regional (NUTS 2) and national.

@@

This study relies on three datasets to analyse RES droughts across the island of Ireland. Data downloaded from C3S-E were used to obtain two datasets: one based on national-level data (C3S-E N), and another on grid-level data (C3S-E G). The third dataset was computed using the Atlite model (Atlite).

@@

For national-level analyses, the aggregated CF time series provided by C3S-E were used at two levels: Republic of Ireland (NUTS0: IE) and Northern Ireland (NUTS2: UKN0). We computed a weighted average of these, based on the installed capacity of each one, to represent the total CF of Ireland. This CF time series for Ireland is based on the underlying C3S-E assumption of RES generation at every 0.25\textdegree grid point in Ireland.

@@

For grid-level analyses, the gridded dataset from C3S-E was used to create a representative time series for Ireland's RES farms. A comprehensive list of RES farms, including each farm’s GPS coordinates and installed capacity, was compiled from the EirGrid website. Using these coordinates, the nearest grid point on the C3S-E grid was identified for each farm. The CF values from the C3S-E dataset corresponding to these grid points were retrieved for each farm. To construct the final CF time series, a weighted average of the CF values was calculated, with the installed capacity of each farm serving as the weight. This process resulted in a single time series for each energy source (wind and PV), providing an accurate representation of RES generation in Ireland at a grid level.

@@

In addition to analysing wind and PV generation separately, a combined CF was computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). Given that PV capacity in Ireland remains relatively low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considered a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These values are based on targets outlined in the roadmap published by EirGrid and SONI~\cite{eirgrid2023future}. New time series were generated for both the Atlite and C3S-E G PV models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remains unchanged, as significant shifts in wind farm distribution are not expected. The PV and wind time series were then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

In total, twelve CF time series were analysed in this study, six for individual wind and PV CF (three models for each source) in the existing scenario, and an additional six time series that include the combined CF for both existing and balanced scenarios across the different models.

@@

In this study, a RES drought event was defined as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period of longer than 24 hours. The choice of this threshold is somewhat arbitrary but aligns with similar studies on low renewable energy production \cite{kaspar2019drought, ohba2022drought, mayer2023drought}. By using a 24-hour moving average, fewer but longer-lasting events were captured compared to using the raw CF time series, which can be more sensitive to short-term fluctuations. A fixed threshold approach was chosen in this study to enable consistent inter-comparison between datasets.

@@

The accuracy of the datasets used in this study was verified, before continuing to the analysis of RES droughts. For the verification process, time-varying values of installed capacity were used to account for changes in RES development over the verification period. This step allowed us to assess how well the datasets represent the production of renewable energy by comparing them against observed data.

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve}a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provides four associated smoothed power curves. The smoothing is done using a Gaussian filter with different standard deviations that depend on the wind speed. A separate wind CF time series for Ireland was generated for each of the wind turbine power curves and smoothing levels. The performance of each CF time series was then assessed based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The Atlite model allows the user to select certain PV panel characteristics. In this study, the three PV panel types available in the Atlite model were considered. Following the same methodology as in the previous section, the three available models were compared using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, the Breyer PV panel model was selected \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity, leading to uncertainties in PV generation data and the actual generating capacity in the first few months after each farm is connected. With over 65\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events.

@@

In this section, RES drought events are evaluated under two different scenarios with fixed installed capacities: the existing scenario, with 89\% of wind capacity (6.7 GW) and 11\% of PV capacity (0.6 GW), and a balanced scenario, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios were driven by 45 years of ERA5 data. Using the RES drought identification process described in Section~\ref{sec:res_drought}, wind and PV droughts are first analysed separately before presenting the results for combined (wind + PV) RES droughts under both scenarios.

@@

For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves differently than the two C3S-E datasets. The Atlite results show a log-linear increase but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets arises from how each model handles PV output near the threshold of 0.1 CF. Atlite remains slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

For the existing installed capacity scenario (Fig.\ref{fig:return_periods}c), the return periods mirror those of wind (Fig.\ref{fig:return_periods}a), due to the low levels of installed PV. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event, shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}, extends from roughly six months for the existing scenario, to four years for the balanced scenario in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets. % The return periods for Atlite align more closely with those of C3S-E in the balanced scenario compared to the existing one. This convergence is due to the reduced proportion of wind capacity in the balanced scenario, which lessens the impact of wind-specific characteristics. Notably, there is a greater difference in how each model represents wind generation than PV, meaning the balanced scenario reduces these wind-related discrepancies, bringing the datasets closer together.

@@

The seasonality of RES droughts was analysed by comparing the percentage of hours in each month classified as part of a RES drought.

@@

In this study, a threshold CF value of 0.1 was used to identify RES droughts. To assess how sensitive the results are to this threshold, the annual number of events was calculated across a range of CF threshold values (Fig.~\ref{fig:number_days_threshold}).

@@

This study highlighted the key differences between the three datasets: Atlite, C3S-E G, and C3S-E N in their ability to model RES droughts. One of the most significant differences is how each dataset incorporates the specific locations of RES farms. Both Atlite and C3S-E G consider the locations of wind and PV farms, which should, in theory, provide a more accurate representation of RES generation. While this approach improves PV drought models, our analysis indicates that for wind, the Atlite dataset performs better overall, especially in its close alignment with EirGrid data for wind generation estimates. This finding suggests that although the inclusion of RES farm locations is beneficial, particularly for PV, the accuracy of the power curve model in Atlite may be the critical factor in producing realistic time series for low-generation events.

The results showed that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. In contrast, Atlite stands out for its closer alignment with EirGrid data, particularly in wind generation, as seen in the verification process (Section \ref{sec:wind_verification}). This distinction was also evident in the analysis, where Atlite reported higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. One of the study's most critical findings is that the accuracy of RES drought modelling relies more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting accurate models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis showed a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. This balanced approach offers a more stable and secure energy supply by mitigating RES drought risks. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.

In terms of future research, there are two main topics to explore. First, future climate data could be integrated with energy scenarios to better understand how climate change might affect RES droughts. Second, expanding the scope of the study to include Europe would provide a more comprehensive understanding of RES droughts in an interconnected energy grid. However, this expansion would require extensive verification across all EU countries, making it a more complex but highly relevant challenge.

\newpage
\section{Acknowledgments}
Using the Climate Data Store. Thank EirGrid for the data and add a reference to the funding project or projects
